The imprint of global magma oceans on exoplanet demographics

### 2023-04-11 src/tex/ms.tex
review Tim's edits of discussion

However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet~\citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples \citep{2018SSRv..214...36A,2022Natur.611..245B}.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,2013NatGe...6..661G,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} \citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen \citep{2013ApJ...778..154W,2014ApJ...785L..20W,2015AsBio..15..119L}.
%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,2022arXiv220310023L}.
Venus' atmospheric composition is consistent with the planet having gone through this phase \citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.

@@

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
%\todo[inline]{TIM: please review} DONE
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}. With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state \citep{2021JGRE..12606711L,2021ApJ...919..130B}. We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean \citep{Dorn2021,2023Icar..39015265S}.

@@

\textit{Super-runaway} planets: For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,2013ApJ...765..131K,Leconte2013,2015ApJ...806..216H,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our sub-runaway planets~\citep{Zeng2016}.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
This scenario does not take into account the expected additional radius increase due to the density deficit of molten interiors~\citep{Bower2019}.

Nominally, the above models assume a \textit{dry melt} without dissolved volatiles.
Here, however, we consider a \textit{wet melt} magma ocean, take into account a radius decrease from water partitioning in the melt, following \citet{Dorn2021}: In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
This removes it from the steam atmosphere, which in turn decreases the radius inflation effect.
The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes are predicted to reach measurable values~\citep{Dorn2021}.
The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
We then add the (in almost all cases negative) radius deviations to the planet radii computed for the dry melt case.

@@

Our finding that the expected trend is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Such constraints will shed light on whether the initial water content of rocky planetary is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects, such as migration \citep{Schlecker2021} and escape \citep{Owen2016}.
Ultimately, such measurements will shed light on the question how delivery and loss effects during rocky planet formation and evolution shape the diversity of exoplanetaty climates.
%Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
%Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
Overall, the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.

@@

\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies sufficient amounts of water vapor available in the atmosphere to act as greenhouse gas, but already about ~10--20 bar of water vapour -- a minor fraction of one Earth ocean (the lower limit of water on Earth) -- is enough to sustain sufficiently high surface temperatures to keep the surface molten, and thus in a magma ocean stage \citep{Boukrouche2021,2021JGRE..12606711L}. The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition. From a formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome of formation \citep{2019PNAS..116.9723Z,2020A&A...643L...1V}. But while ``typically'' considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System \citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation. Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amount of water inside of the planet during the magma ocean phase \citep{2018SSRv..214...76I,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}. Enhanced equilibration between the core, mantle, and atmospehre during magma ocean evolution of rocky exoplanets further enhances this process \citep{2021ApJ...914L...4L,Schlichting2022}. Therefore, even nominally dry planets generate substantial amounts of water during formation and early evolution. Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M dwarf systems \citep{2022ApJ...938L...3L}.

\textit{Planetary evolution and duration of the steam atmosphere phase}: An inflated steam atmosphere can only be sustained until the planet has lost its water. Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal. However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of oceans \citep{2018AJ....155..195W} on even the most irradiated exoplanets \citep[see the discussion in][]{2022ApJ...938L...3L}.

\textit{Prevalent water inventory}: The magnitude of radius change at the runaway greenhouse threshold is sensitive to the water mass fraction. As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern: The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection. Howver, dissolution into the magma ocean and atmospheric inflation show non-linear coupling; thus a straightforward trend is difficult to discern.

@@

\subsubsection{False positive scenarios}\label{sec:dis_falsepositive}
Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{2012ApJ...753...66I,Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates. For example, the K2-3 system shows tentative evidence for a break in the planet radii, contrary to what would be expected from atmospheric escape \citep{2022AJ....164..172D}.

@@

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{2020A&A...643L...1V,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals \citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.

@@

We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population. (i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared. In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m \citep[e.g.][]{2021ApJ...919..130B}. This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE \citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}. Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime \citep{2020SSRv..216...98G}. (ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss \citep{2014ApJ...785L..20W,2015AsBio..15..119L}. (iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere \citep{2019JGRE..124.2015K,2021JGRE..12606643K}. This may be detectable in high-resolution observations focusing on isotope-sensitive transitions \citep{2019AJ....158...26L,2019A&A...622A.139M}. (iv) Disequilibrium chemistry in tidally-locked super-runaway planets. It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers \citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface \citep{2019ApJ...887..231L,2021ApJ...921L...8H}. Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma \citep{2022E&PSL.59817847G,2022PSJ.....3...93B}. Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed. This has been suggested to be done via (v) eclipse photometry \citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean \citep{2020ApJ...898..160E,Fortin2022}.

\subsubsection{Detection in multi-planet systems}
As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, K2-3 \citep{2022AJ....164..172D}, or Kepler-138 \citep{2022NatAs.tmp..269P}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
The current sample of suitable systems is sparse: The California-Kepler Survey catalog~\citep{Fulton2018} contains only six planets with instellations $< 2\, S_\oplus$ and smaller than \SI{2}{\rEarth} in five multi-planet systems.
Future additions to the multi-planet sample through missions such as \plato\ are needed.

@@

Other mission concepts exist that focus on characterizing exoplanet habitability, most notably the \hwo~\citep{Gaudi2020c,LUVOIR2019} and the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}.
In principle, direct imaging surveys do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets. However, because mid-infrared retrievals do not suffer from the degeneracy between cloud albedo and changes in surface area, the planet radius can be retrieved in mid-infrared wavelengths \citep{2018ExA....46..543D,2021ExA...tmp..118Q}. Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light \citep{Wordsworth2022}. Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios (Konrad+, Alei+). Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages \citep{2014ApJ...784...27L,2019A&A...621A.125B}. With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states \citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.

@@

\subsubsection{Baseline mass-radius relationship}
Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure \citep[][, add Noack+ paper]{2016ApJ...819..127Z,2021JGRE..12606724B}.
We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
Nevertheless, we performed a sanity check to assess the influence varying our baseline model on our results (see Appendix~\ref{app:MR_relation}), and found general agreement between different interior composition models.

>>>>

However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet~\citep{Hamano2013}, which is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,2013NatGe...6..661G,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,2015AsBio..15..119L}.
%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,2022arXiv220310023L}.
Venus' atmospheric composition is consistent with the planet having gone through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.

@@

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{2021JGRE..12606711L,2021ApJ...919..130B}.
We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,2023Icar..39015265S}.

@@

For \textit{super-runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,2013ApJ...765..131K,Leconte2013,2015ApJ...806..216H,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our sub-runaway planets~\citep{Zeng2016}.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
This scenario does not take into account the expected additional radius increase due to the density deficit of molten interiors~\citep{Bower2019}.

Nominally, the above models assume a \textit{dry melt} without dissolved volatiles.
Here, however, we consider a \textit{wet melt} magma ocean and take into account a radius decrease from water partitioning in the melt~\citep{Dorn2021}: In the situation of a magma ocean below a steam atmosphere, part of the water will partition into the melt.
This removes it from the steam atmosphere, which in turn decreases the radius inflation effect.
The effect is generally small compared to the radius inflation from a steam atmosphere but resulting radius changes are predicted to reach measurable values~\citep{Dorn2021}.
The magnitude of this effect is again dependent on the planet mass and on the water mass fraction.
We follow \citet{Dorn2021} and use their computed radius deviations between a wet magma ocean and a solid mantle for a tropopause pressure~$P_\mathrm{iso}=\SI{0.1}{\bar}$.
We then add the (in almost all cases negative) radius deviations to the planet radii computed for the dry melt case.

@@

Our finding that the expected trend is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Such constraints will provide insight into whether the initial water content of rocky planets is bifurcated by systematic inter-system effects~\citep{Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects such as planet migration~\citep{Schlecker2021} or atmospheric escape~\citep{Owen2016}.
Ultimately, such measurements will shed light on how delivery and loss effects during rocky planet formation and evolution shape the diversity of exoplanetary climates.
%Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
%Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
Overall, probing the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.

@@

\textit{Occurrence rate of planets forming steam atmospheres}: The runaway greenhouse climate relies on sufficient amounts of atmospheric water vapor that can act as a greenhouse gas.
However, already about \SIrange{\sim 10}{20}~bar of water vapor -- corresponding to a minor fraction of one Earth ocean and thus the lower limit of water on Earth -- is enough to sustain sufficiently high surface temperatures to keep the planet in a magma ocean stage~\citep{Boukrouche2021,2021JGRE..12606711L}.
The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
From a planet formation perspective, the incorporation of water into planets in the terrestrial planet zone is a standard outcome~\citep{2019PNAS..116.9723Z,2020A&A...643L...1V}.
But while commonly considered volatile delivery channels suggest a fraction of planets to be volatile-poor, the incorporation of hydrogen into even the driest planetary materials known in the Solar System~\citep{2020Sci...369.1110P,2021PSJ.....2..244J} suggests that hydrogen is present in all rocky planets upon formation.
Accreted hydrogen in nominally dry planetary materials react with mantle oxygen to form substantial amounts of water inside of the planet during the magma ocean phase~\citep{2018SSRv..214...76I,2021ApJ...909L..22K,2020MNRAS.496.3755K,2022NatAs...6.1296K}.
Enhanced equilibration between the core, mantle, and atmosphere during magma ocean evolution of rocky exoplanets further enhances this process~\citep{2021ApJ...914L...4L,Schlichting2022}.
Therefore, even nominally dry planets generate substantial amounts of water during formation and early evolution.
Reduced heating from short-lived radionuclides in extrasolar planetary systems increases the expected water abundance in exoplanet systems further, in particular for M~dwarf systems~\citep{2022ApJ...938L...3L}.

\textit{Planetary evolution and duration of the steam atmosphere phase}: An inflated steam atmosphere can only be sustained until the planet has lost its water.
Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of terrestrial oceans per Gyr~\citep{2018AJ....155..195W} on even the most irradiated exoplanets~\citep[see discussion in][]{2022ApJ...938L...3L}.

\textit{Prevalent water inventory}: The magnitude of radius change at the runaway greenhouse threshold is sensitive to the water mass fraction.
As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern:
The higher the water content and the higher the fraction of planets in runaway greenhouse climates, the greater the prospect of a statistically sound detection.
However, dissolution into the magma ocean and atmospheric inflation show non-linear coupling; thus, a straightforward trend is difficult to discern.

@@

\subsubsection{False positive scenarios}\label{sec:dis_falsepositive}
Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{2012ApJ...753...66I,Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.
For example, the \mbox{K2-3} system shows tentative evidence for a break in the planet radii, contrary to what would be expected from atmospheric escape~\citep{2022AJ....164..172D}.

@@

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
However, a population of ``water worlds'' with low bulk densities on a wide range of orbits would dilute the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{Venturini2020,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals~\citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this dilution.

@@

We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population. (i) Detecting the atmospheric windows of water vapour in the near- to mid-infrared. In a super-runaway state the atmosphere is dominated by the opacity of water vapour, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.][]{2021ApJ...919..130B}. This would require an instrument with capabilities in this wavelenght regime, for instance ELT METIS, JWST MIRI, or future missions such as LIFE~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}. Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}. (ii) Post-runaway planets may be detectable by O$_{3}$ absoprtion due to build-up of abiotic oxygen, left-over from photochemical dissociation and hydrogen loss~\citep{2014ApJ...785L..20W,2015AsBio..15..119L}. (iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}. This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}. (iv) Disequilibrium chemistry in tidally-locked super-runaway planets. It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}. Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}. Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed. This has been suggested to be done via (v) eclipse photometry~\citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidifed magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.

\subsubsection{Detection in multi-planet systems}
As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, K2-3~\citep{2022AJ....164..172D}, or Kepler-138~\citep{2022NatAs.tmp..269P}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
The current sample of suitable systems is sparse: The California-Kepler Survey catalog~\citep{Fulton2018} contains only six planets with instellations $< 2\, S_\oplus$ and smaller than \SI{2}{\rEarth} in five multi-planet systems.
Future additions to the multi-planet sample through missions such as \plato\ are needed.

@@

Other mission concepts exist that focus on characterizing exoplanet habitability, most notably the \hwo~\citep{Gaudi2020c,LUVOIR2019} and the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2021}.
In principle, direct imaging surveys do not directly measure planetary radii and are therefore most useful for providing context through atmospheric measurements of individual planets. However, because mid-infrared retrievals do not suffer from the degeneracy between cloud albedo and changes in surface area, the planet radius can be retrieved in mid-infrared wavelengths~\citep{2018ExA....46..543D,2021ExA...tmp..118Q}. Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}. Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios (Konrad+, Alei+). Finally, since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}. With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states \citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.

@@

\subsubsection{Baseline mass-radius relationship}
Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep[][, add Noack+ paper]{2016ApJ...819..127Z,2021JGRE..12606724B}.
We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
Nevertheless, we performed a sanity check to assess the influence varying our baseline model on our results (see Appendix~\ref{app:MR_relation}), and found general agreement between different interior composition models.
